Untangling the link between traits, size and growth rate in plants

### 2015-02-18 ms.tex
Enhance introduction

Functional traits are thought to capture core differences in the strategies
plants use to generate and invest surplus energy
\citep{wright_world-wide_2004,chave_towards_2009,westoby_plant_2002}. Useful
traits are those that
can be easily measured for diverse species and where differences between
species are large compared to intra-specific variation. Moroever, the uptake
of traits reflects the observation that although most plants have same basic
construction, physiological function and resource requirements, species differ
considerably from one another in rates of carbon, nitrogen and water uptake,
growth, mortality and seed production. Data for prominent traits now exists
for many of the world's 250000 plant species \citep{cornwell_functional_2014}.
The effect of some traits of multiple aspects of plant function has also been
well documented, including mechanical strength, biomass partitioning, rates of
photosynthesis and rates of tissue turnover
\citep {wright_world-wide_2004,chave_towards_2009}. As a result, the idea has
taken hold that the growth startegy and ecology of a plant species might usefully
be characterised simply by knowing its traits.

While the influence of plant traits on elements of plant physiologival
function is increasingly understood, attempts at using traits to predict
demographic rates have only met with mixed success. In seedlings, the trait
leaf mass per area (LMA) -- a central part of the leaf economics spectrum --
is tightly related to plant relative growth rate in mass. LMA and its
correlate leaf lifespan have also been linked to height growth rate for small
seedlings and saplings. These early successes prompted researchers to search
for similar relationships in large plants. However, the results showed that in
large plants, LMA was not correlated with diameter growth rate. Meanwhile,
other traits such as wood density showed strong relationships
to growth in large plants, but only weak relationships in seedlings \citep
{castro-diez_stem_1998}. To date each relationship has been quantified mostly for
one or few traits and at a single life-stage. Yet the picture is emerging is that the
link between traits and growth may be modified by plant size; a finding which would
challenge the asummption applied by many that particular ends of the trait spectrum
translate directly into fast or slower growth.

The challenge of interpreting diverse empirical results linking traits to
growth rate -- or not -- would clearly be easier if we had clearl expectations
on what signal one should expect, given our understanding of how traits
infleunce plant function. This is a primary, yet perhaps underappreciated,
role for theory: to lay expectations based on mechistic principles against
which empircal data can be compared. Current empircial results suggest the
effect of traits on growth changes with size and possibly also light
environment, yet current theory says little about how relationships might come
about. A widely-used model suggests mass-based relative growth rate will be a
linear function of LMA, at least for seeldings.  An extension of model
suggests a similar relationship may hold at larger sizes. But this prediction
doesn't empirical results. Meanwhile, theoretical predictions on how other
traits should influence growth are largely absent.

A further problem with existing models is that the effects of traits are all
realised via mass production, whereas the effect of some prominent traits is
most likely on the quality of tissue construction, or the amoutn of mass
allocated to different tissues. There are two problem here. First, measuring mass
production is only practical for small plants that can be harvested. For larger plants,
growth is more often emasured either as diameter or height growth. For example, the
CTFS monitors diameter growth in 49 large permenenat plots worldwide. Establishing
the link between traits, mass growth and diamater growth is thus essential for predicting
the effect of traits on growth across these diverse datasets.

Here we provide a mechanistic framework describing how traits link to
demography, linking to commonly used metrics such as height and diameter
growth, combine this with fresh empirical insight using largest
collation of growth and trait data available, from long-term plot at
BCI, Panama.

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Shows why effect of traits changes with size,size dependent effect
  differs among traits.

  \begin{itemize}
  \itemsep1pt\parskip0pt\parsep0pt
  \item
    Contradicts widespread assumption that traits relate to growth in
    simple linear way
  \end{itemize}
\item
  Offers framework for interpreting and understanding empirical results
\end{itemize}

\section*{Fresh empirical motivation}\label{fresh-empirical-motivation}

To help motivate our model, we begin with a fresh analysis of growth dynamics
in the long-term plot at Barro Colorado Island in Panam, where the growth
rates for over 100 species have been recorrded over the last 30 years. A
prior analysis by \citet{wright_functional_2010} found a mdoerate effect of
wood density on growth rate, but almsot no effect of LMA, and only a weak
effect of a thrid trait maximum plant height (HMAX).

Fresh empirical analysis highlights how effect of traits on pot growth
changes with size.

@@

  \begin{itemize}
  \itemsep1pt\parskip0pt\parsep0pt
  \item
    details about our model ()
  \item
    show argument hold intuiatively
  \end{itemize}
\end{itemize}

>>>>

Functional traits are thought to capture core differences in the strategies
plants use to generate and invest surplus energy
\citep{wright_world_wide_2004, chave_towards_2009, westoby_plant_2002}.
Although most plants have the same basic construction, physiological function
and resource requirements, species differ considerably from one another in
rates of carbon, nitrogen and water uptake, growth, mortality and seed
production. Some of these functional differences have been linked to
underlying variation in relatively simple physiological traits, such as leaf
mass per area (LMA), wood density (WD), leaf nitrogen per area, vessel
diameter,  and seed size  \citep{wright_world_wide_2004,chave_towards_2009}.
These traits are easily measured across diverse species, and variation between
species is large compared to variation within species. Data for some traits
now exists for up to 10\% of the world's 250000 plant species
\citep{cornwell_functional_2014}. During the last decade, the idea has taken
hold that the growth strategy and ecology of a plant species might usefully be
characterised simply by knowing its traits.

While the influence of plant traits on elements of plant physiological
function is increasingly understood, attempts at using traits to understand
demographic rates have only met with mixed success
\citep{wright_functional_2010, poorter_are_2008}. In seedlings, the trait LMA
-- a central part of the leaf economics spectrum
\citep{wright_world_wide_2004} --is tightly related to plant relative growth
rate in mass \citep{lambers_inherent_1992, wright_cross_species_2000}. LMA and
its correlate leaf lifespan have also been linked to height growth rate for
small seedlings and saplings \citep{reich_leaf_1992, poorter_leaf_2006}. These
early successes prompted researchers to search for similar relationships in
large plants. However, the results showed that in trees LMA was not correlated
with diameter growth rate  \citep{wright_functional_2010, poorter_are_2008,
herault_functional_2011}. Meanwhile, other traits such as wood density showed
strong relationships to growth in large plants \citep{wright_functional_2010},
but only weak relationships in seedlings  \citep{castro-diez_stem_1998}. Thus
far, most trait growth relationships have been quantified for one or few
traits and at a single life- stage. Yet the picture is emerging is that the
link between traits and growth may be modified by plant size
\citep{ruger_functional_2012}; a finding that challenges the wide-spread
assumption that particular ends of the trait spectrum translate directly into
faster or slower growth.

The challenge of interpreting diverse empirical results linking  -- or not --
traits to growth rate would clearly be easier with clear expectations on what
signal one should expect, given current understandings of how traits influence
plant function. Generating these expectations is a primary, yet perhaps under-
appreciated, role for theory \citep{kokko_modelling_2007}. Current empirical
results suggest the effect of traits on growth changes with size and possibly
also light environment, yet current theory says little about how such
relationships might come about. A widely-used model shows that for seedlings,
mass-based relative growth rate is linearly and negatively related to LMA
\citep{lambers_inherent_1992, cornelissen_seedling_1996,
wright_cross_species_2000}. An extension of the model suggests a similar
relationship may hold at larger sizes \citep{enquist_general_2007}. But as
described above, this prediction is not supported by empirical results.
Meanwhile, theoretical predictions on how other traits should influence growth
are largely absent.

A further problem with existing theory is that the effects of traits are all
realised via influences on mass production\citep{enquist_general_2007},
whereas the effect of traits such as LMA, WD, and Hmax is via the quality of
tissue construction, or the amount of mass allocated to different tissues
\citep{falster_influence_2011}. There are in fact two problems with theory
focussed solely on mass production. The first is that measuring mass
production is only practical for small plants that can be easily harvested.
For larger plants, growth is more often measured either as diameter or height
growth. For example, the CTFS monitors diameter growth in 49 large permanent
plots across forests worldwide \citep{anderson_teixeira_ctfs_forestgeo_2014},
while forestry agencies around the world have likewise recorded data across
large tracts of forest \citep{purves_predictive_2008}. The second problem is
that, as we outline further below, because mass production is only one
component of plant growth; theory focussing solely on mass production will
therefore always be limited.

In this paper we provide a mechanistic framework for describing the potential
effect of traits on growth throughout a plant's life-cycle, linking to
commonly used metrics such as height and diameter growth. To help motivate our
study we re-analyse demographic and trait data from an earlier study
\citep{wright_functional_2010}, to show 1) that the link between traits and
growth is stronger than previously thought, 2) that the influence of three
prominent traits on growth rate is moderated by plant size, and 3) that the
impact of size on the nature of relationships varies among traits (see also
\citet{ruger_functional_2012}. We then show how these results align with new
set of theoretical predictions. Moreover, we show how traits moderate plant
responses to light environment and also determine shade tolerance, supporting
diverse empirical results \citep{ruger_functional_2012, poorter_leaf_2006}. By
disentangling the effects of plant size, light environment and traits on
growth rates, we hope to provide a solid theoretical foundation for trait
ecology and thus provide a platform for understanding growth across diverse
species around the world.

\subsection*{Empirical motivation}\label{fresh-empirical-motivation}

To help illustrate the empirical patterns, we begin with a fresh analysis of
growth dynamics in the long-term plot at Barro Colorado Island in Panama,
where the growth rates for over 100 species have been recorded over the last
30 years. Data from a long term plot \citep{condit_barro_2012} are combined
with results of experiment on seedlings \citep{kitajima_leaf_2013}, to better understand
size-related patterns of growth. A prior analysis by
\citet{wright_functional_2010} found a moderate effect of wood density on
growth rate, but almost no effect of LMA, and only a weak effect of HMAX. More recently,
\citet{ruger_functional_2012} showed

@@

  \begin{itemize}
  \itemsep1pt\parskip0pt\parsep0pt
  \item
    details about our model ()
  \item
    show argument hold intuitively
  \end{itemize}
\end{itemize}

### 2015-02-18 notes/ms_short.tex
Enhance introduction

Plant growth is the foundation of the terrestrial biosphere, powering
food-webs, the carbon cycle and much of human
activity \citep{raupach_carbon_2010}. Plant species differ in many
functional traits, including aspects of tissue construction and energy
allocation, that drive differences in mechanical strength, biomass
partitioning, and rates of photosynthesis and tissue
turnover\citep{wright_world-wide_2004,chave_towards_2009, westoby_plant_2002}. Yet, it remains unclear how -- and even if --
such traits influence whole-plant growth, with the simple relationships
predicted by existing theory often lacking empirical
support\citep{wright_functional_2010,poorter_are_2008}. Here we
provide a new set of theoretical predictions, validated using 20 years
of growth data\citep{condit_barro_2012,kitajima_leaf_2013}, showing
that the influence of three prominent traits on growth rate is moderated
by plant size, and that the impact of size on the nature of
relationships varies among traits. We extend a widely-used theoretical
model that links growth rate in seedlings with a single leaf
trait\citep{lambers_inherent_1992,wright_cross-species_2000} to
explicitly include influences of size, light environment, and three
other prominent traits. We find the influence of leaf construction cost
(\(\phi\)) on growth is strong in small plants but weakens with size. In
contrast, the influence of stem construction cost (\(\rho\)) peaks at
intermediate sizes, while the influence of height at maturation
(\(h_m\)) strengthens with size. Our theoretical model suggests the
change in trait-growth relationships with size arises due to changes in
biomass partitioning and reproductive allocation. Moreover, we show how
traits moderate plant responses to light environment and also determine
shade tolerance, supporting diverse empirical
results\citep{ruger_functional_2012,poorter_leaf_2006}. By
disentangling the effects of plant size, light environment and traits on
growth rates, our results provide a solid theoretical foundation for
trait ecology and thus provide a platform for understanding growth
across diverse species around the world.

@@

Functional traits capture core differences in the strategies plants use
to generate and invest surplus energy\citep{wright_world-wide_2004, chave_towards_2009,westoby_plant_2002}. Data for prominent
traits now exists for many of the world's 250000 plant
species\citep{cornwell_functional_2014}; yet the effect of these traits
on growth remains poorly understood. Prominent traits such as \(\phi\)
(also known as leaf mass per area) are widely considered as indicators
of growth rate\citep{reich_world-wide_2014}, because cheaper leaf
construction should enable faster canopy
expansion\citep{blackman_compound_1919,westoby_time_2000}. Indeed,
growth rate in seedlings is strongly and negatively correlated with
\(\phi\)\citep{cornelissen_seedling_1996,wright_cross-species_2000}
(Fig. \ref{f-BCI}). Although, it has been suggested that low \(\phi\)
might also give a growth advantage in larger
plants\citep{enquist_general_2007}, an increasing number of empirical
studies have failed to detect any such
effect\citep{wright_functional_2010, @herault_functional_2011,
@poorter_are_2008}. Meanwhile, expectations on how other prominent
traits might influence growth are lacking. The growing discord between
theoretical expectation and empirical result thus indicates a strong
need for a new theoretical framework through which the influence of
traits and size on plant growth can be understood.

Here we extend a widely-used theoretical model that links growth rate in
seedlings to \(\phi\)\citep{lambers_inherent_1992, wright_cross-species_2000}, to explicitly include influences of
size, light environment, and other prominent traits. This extension
allows us to generate novel predictions on the link between four key
traits and plant demography (Table \ref{tab:predictions}). Predictions
are compared against data from both a long-term
plot\citep{condit_barro_2012} and field
experiment\citep{kitajima_leaf_2013} at Barro Colorado Island, Panama.
We use a new method to estimate species potential growth rate,
accounting for size specific effects (Fig. \ref{S-quantile_examples},
see Methods for details). This new method allowed us to extract a more
definite signal of trait effects than in previous
studies\citep{wright_functional_2010}. The results of this new analysis
(Fig. \ref{f-BCI}) align closely with our theoretical predictions (Table
\ref{tab:predictions}) showing how the effect of traits on growth
changes with plant size, with different traits exhibiting distinct
size-dependent effects.

@@

Although Eq. \ref{eq:RGR} captures patterns of growth in seedlings in
relation to \(\phi\)\citep{wright_cross-species_2000}, this
approximation does not directly link to other traits, or to the
variables that are routinely collected for large trees: namely plant
height (\(h\)) and stem cross-sectional area (\(a_\textrm{st}\)) or
diameter \(D\).

@@

By definition, the main effect of increasing \(\phi\) is to decrease the
marginal leaf deployment per mass (eq. \ref{eq:daldmt}) and thus reduce
growth rate. However, to fully account for the influence of \(\phi\) on
growth we must also account for the decrease in leaf turnover that
results from superior leaf construction\citep{wright_world-wide_2004}.
This is achieved by linking the turnover rate of leaf in eq.
\ref{eq:dPdt} to \(\phi\). We use an allometric scaling relationship of
the form \(k_\textrm{l}=\alpha_4 \, \phi^{-\beta_4}\), which has been
shown to describe patterns across diverse
ecosystems\citep{wright_world-wide_2004} (Fig. \ref{fS-leaf}).

@@

\begin{figure}[htbp]
\centering
\includegraphics{figures/SI_lma_tradeoff.pdf}
\caption{\textbf{Leaf turnover decreases with leaf-construction cost.}
Data from \citep{wright_world-wide_2004} for 678 species from 51 sites, each
point giving a species-average. Lines show standardised major axis lines
fitted to data from each site, with intensity of shading adjusted
according to strength of the relationship.\label{fS-leaf}}
\end{figure}

>>>>

Plant growth is the foundation of the terrestrial biosphere, powering
food-webs, the carbon cycle and much of human
activity \citep{raupach_carbon_2010}. Plant species differ in many
functional traits, including aspects of tissue construction and energy
allocation, that drive differences in mechanical strength, biomass
partitioning, and rates of photosynthesis and tissue
turnover\citep{wright_world_wide_2004,chave_towards_2009, westoby_plant_2002}. Yet, it remains unclear how -- and even if --
such traits influence whole-plant growth, with the simple relationships
predicted by existing theory often lacking empirical
support\citep{wright_functional_2010,poorter_are_2008}. Here we
provide a new set of theoretical predictions, validated using 20 years
of growth data\citep{condit_barro_2012,kitajima_leaf_2013}, showing
that the influence of three prominent traits on growth rate is moderated
by plant size, and that the impact of size on the nature of
relationships varies among traits. We extend a widely-used theoretical
model that links growth rate in seedlings with a single leaf
trait\citep{lambers_inherent_1992,wright_cross_species_2000} to
explicitly include influences of size, light environment, and three
other prominent traits. We find the influence of leaf construction cost
(\(\phi\)) on growth is strong in small plants but weakens with size. In
contrast, the influence of stem construction cost (\(\rho\)) peaks at
intermediate sizes, while the influence of height at maturation
(\(h_m\)) strengthens with size. Our theoretical model suggests the
change in trait-growth relationships with size arises due to changes in
biomass partitioning and reproductive allocation. Moreover, we show how
traits moderate plant responses to light environment and also determine
shade tolerance, supporting diverse empirical
results\citep{ruger_functional_2012,poorter_leaf_2006}. By
disentangling the effects of plant size, light environment and traits on
growth rates, our results provide a solid theoretical foundation for
trait ecology and thus provide a platform for understanding growth
across diverse species around the world.

@@

Functional traits capture core differences in the strategies plants use
to generate and invest surplus energy\citep{wright_world_wide_2004, chave_towards_2009,westoby_plant_2002}. Data for prominent
traits now exists for many of the world's 250000 plant
species\citep{cornwell_functional_2014}; yet the effect of these traits
on growth remains poorly understood. Prominent traits such as \(\phi\)
(also known as leaf mass per area) are widely considered as indicators
of growth rate\citep{reich_world_wide_2014}, because cheaper leaf
construction should enable faster canopy
expansion\citep{blackman_compound_1919,westoby_time_2000}. Indeed,
growth rate in seedlings is strongly and negatively correlated with
\(\phi\)\citep{cornelissen_seedling_1996,wright_cross_species_2000}
(Fig. \ref{f-BCI}). Although, it has been suggested that low \(\phi\)
might also give a growth advantage in larger
plants\citep{enquist_general_2007}, an increasing number of empirical
studies have failed to detect any such
effect\citep{wright_functional_2010, @herault_functional_2011,
@poorter_are_2008}. Meanwhile, expectations on how other prominent
traits might influence growth are lacking. The growing discord between
theoretical expectation and empirical result thus indicates a strong
need for a new theoretical framework through which the influence of
traits and size on plant growth can be understood.

Here we extend a widely-used theoretical model that links growth rate in
seedlings to \(\phi\)\citep{lambers_inherent_1992, wright_cross_species_2000}, to explicitly include influences of
size, light environment, and other prominent traits. This extension
allows us to generate novel predictions on the link between four key
traits and plant demography (Table \ref{tab:predictions}). Predictions
are compared against data from both a long-term
plot\citep{condit_barro_2012} and field
experiment\citep{kitajima_leaf_2013} at Barro Colorado Island, Panama.
We use a new method to estimate species potential growth rate,
accounting for size specific effects (Fig. \ref{S-quantile_examples},
see Methods for details). This new method allowed us to extract a more
definite signal of trait effects than in previous
studies\citep{wright_functional_2010}. The results of this new analysis
(Fig. \ref{f-BCI}) align closely with our theoretical predictions (Table
\ref{tab:predictions}) showing how the effect of traits on growth
changes with plant size, with different traits exhibiting distinct
size-dependent effects.

@@

Although Eq. \ref{eq:RGR} captures patterns of growth in seedlings in
relation to \(\phi\)\citep{wright_cross_species_2000}, this
approximation does not directly link to other traits, or to the
variables that are routinely collected for large trees: namely plant
height (\(h\)) and stem cross-sectional area (\(a_\textrm{st}\)) or
diameter \(D\).

@@

By definition, the main effect of increasing \(\phi\) is to decrease the
marginal leaf deployment per mass (eq. \ref{eq:daldmt}) and thus reduce
growth rate. However, to fully account for the influence of \(\phi\) on
growth we must also account for the decrease in leaf turnover that
results from superior leaf construction\citep{wright_world_wide_2004}.
This is achieved by linking the turnover rate of leaf in eq.
\ref{eq:dPdt} to \(\phi\). We use an allometric scaling relationship of
the form \(k_\textrm{l}=\alpha_4 \, \phi^{-\beta_4}\), which has been
shown to describe patterns across diverse
ecosystems\citep{wright_world_wide_2004} (Fig. \ref{fS-leaf}).

@@

\begin{figure}[htbp]
\centering
\includegraphics{figures/SI_lma_tradeoff.pdf}
\caption{\textbf{Leaf turnover decreases with leaf-construction cost.}
Data from \citep{wright_world_wide_2004} for 678 species from 51 sites, each
point giving a species-average. Lines show standardised major axis lines
fitted to data from each site, with intensity of shading adjusted
according to strength of the relationship.\label{fS-leaf}}
\end{figure}
